---
title: Hyper CGI methylation visualization
output: pdf_document
params:
  cgi_path: ""
  sample_path: ""
---

<!--
// ==========================================================================
//                 Cancer model DNA methylation visualization               
// ==========================================================================
// Copyright (c) 2025, Sara Hetzel <hetzel @ molgen.mpg.de>
// Copyright (c) 2025, Max-Planck-Institut für Molekulare Genetik
// All rights reserved.
//
// This script is licensed under the BSD 3-Clause License.
// See the LICENSE file distributed with this source for full terms.
//
// This script is distributed in the hope that it will be useful,
// but WITHOUT ANY WARRANTY; without even the implied warranty of
// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.
// ==========================================================================
// 
// ==========================================================================
-->

```{r echo=FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(knitr)
```

# Assessing CGI methylation of cancer models and primary tumors

```{r echo=FALSE, warning = FALSE}
## Parse arguments

if (params$cgi_path == "") stop("Please provide path to CGI averages.")
if (params$sample_path == "") stop("Please provide sample file in csv format.")
```

```{r echo=FALSE, warning = TRUE}
################################################################################
## Load data objects
################################################################################

load(file = "data-objects/prc2_cgis_human.RData")
load(file = "data-objects/prc2_cgis_mouse.RData")
load(file = "data-objects/metadata_human.RData")
load(file = "data-objects/metadata_mouse_models.RData")
load(file = "data-objects/common_hyper_cgi_across_tumor_types_array_any_purity.RData")
load(file = "data-objects/hyper_cgi_per_tumor_type_array_any_purity.RData")
load(file = "data-objects/cgi_hg19_mm10_mapping.RData")

################################################################################
## Define CGI sets
################################################################################

hyper_pancancer_human <- common_hyper_cgi_across_tumor_types_array_any_purity
hyper_pancancer_mouse <- cgi_hg19_mm10_mapping$pancancer

################################################################################
## Define colors and shapes
################################################################################

colors_condition <- c("#979EA4", "#D1D3D4", "#7ACDCD", "#2FA19E")
names(colors_condition) <- c("healthy", "healthy_sorted", "precursor", "tumor")

shapes_condition <- c(19, 1, 2, 17)
names(shapes_condition) <- c("healthy", "healthy_sorted", "precursor", "tumor")

shapes_group <- c(19, 4)
names(shapes_group) <- c("primary", "model")

colors_type <- c("#3F007D", "#54278F", "#6A51A3", "#807DBA", "#9E9AC8", "#BCBDDC", "#DADAEB", "#7A0177", "#AE017E", "#F768A1", "#FA9FB5", "#FCC5C0", "#FDE0DD", "#8C2D04", "#CC4C02", "#EC7014", "#FE9929", "#FEC44F", "#FEE391", "#08306B", "#08519C", "#2171B5", "#4292C6", "#6BAED6", "#9ECAE1", "#C6DBEF")
names(colors_type) <- c("T-ALL", "CESC", "B-ALL", "ESCA", "CHOL", "HNSC", "COAD", "PRAD", "BRCA", "PAAD", "GBM", "LUAD", "READ", "BLCA", "UCEC", "LUSC", "LAML", "SKCM", "LIHC", "STAD", "KIRC", "SARC", "PCPG", "KIRP", "THYM", "THCA")

################################################################################
## Load new samples
################################################################################

sample_sheet <- read.csv(params$sample_path, header = TRUE, stringsAsFactors = FALSE, row.names = 1)

samples_human <- rownames(subset(sample_sheet, species == "human"))
samples_mouse <- rownames(subset(sample_sheet, species == "mouse"))

if (any(!sample_sheet$type %in% names(hyper_cgi_per_tumor_type_array_any_purity))) warning("One or more tumor types do not have a pre-defined hyper CGI set. Only the PRC2 and the pan-cancer hyper CGI set will be shown.")
if (any(!sample_sheet$condition %in% c("healthy", "healthy_sorted", "precursor", "tumor"))) stop("Condition must be one of: 'healthy', 'healthy_sorted', 'precursor', 'tumor'.")
if (any(!sample_sheet$group %in% c("primary", "model"))) stop("Group must be one of: 'primary', 'model'.")

```

Calculating metrics for `r  length(samples_human)` new human and `r  length(samples_mouse)` new mouse samples.

Tumor types included in new data: `r  paste(unique(sort(sample_sheet$type)), sep = ", ")`.

New sample summary per group, species and condition:

```{r echo=FALSE, warning = FALSE}

df_summary <- data.frame(table(sample_sheet[,c("condition", "species", "group")]))
df_summary <- subset(df_summary, Freq != 0)
colnames(df_summary) <- c("condition", "species", "group", "count")
rownames(df_summary) <- NULL
df_summary <- df_summary[,c("group", "species", "condition", "count")]
kable(df_summary)
```

```{=latex}
\newpage
```

```{r echo=FALSE, warning = FALSE}
## Human
cgi_files <- list.files(params$cgi_path, "avg_cgi", full.names = TRUE)
avg_cgi <- lapply(cgi_files, read.table, stringsAsFactors = FALSE)
names(avg_cgi) <- sapply(cgi_files, function(x) strsplit(basename(x), "avg_cgi_|\\.bed")[[1]][2])

if (!all(rownames(sample_sheet) %in% names(avg_cgi))) stop("Sample names differ between sample sheet and CGI files.")

## Load and merge CGIs - human

df_cgi_summary_human <- data.frame(row.names = samples_human, type = sample_sheet[samples_human,"type"], condition = sample_sheet[samples_human,"condition"], group = sample_sheet[samples_human,"group"])

if (length(samples_human) > 0)
{
    avg_cgi_human <- lapply(samples_human, function(x)
    {
        data <- avg_cgi[[x]]
        colnames(data) <- c("chr", "start", "end", x)
        return(data)
    })

    avg_cgi_human <- Reduce(function(x,y) merge(x = x, y = y, all.x = TRUE, all.y = TRUE), avg_cgi_human)
    rownames(avg_cgi_human) <- paste(avg_cgi_human[,1], avg_cgi_human[,2], avg_cgi_human[,3], sep = "-")
    avg_cgi_human <- avg_cgi_human[,-c(1:3)]

    ## Fraction methylated
    df_cgi_summary_human$class <- "new"

    df_cgi_summary_human$fraction_meth_prc2 <- apply(avg_cgi_human[rownames(avg_cgi_human) %in% prc2_cgis_human,samples_human], 2, function(x) length(which(x > 0.2)) / length(na.omit(x)))
    df_cgi_summary_human$fraction_meth_pancancer <- apply(avg_cgi_human[rownames(avg_cgi_human) %in% hyper_pancancer_human,samples_human], 2, function(x) length(which(x > 0.2)) / length(na.omit(x)))
    df_cgi_summary_human$fraction_meth_tumor_type <- sapply(samples_human, function(x)
    {
        curr_type <- df_cgi_summary_human[x,"type"]
        curr_data <- avg_cgi_human[rownames(avg_cgi_human) %in% hyper_cgi_per_tumor_type_array_any_purity[[curr_type]],x]
        return(length(which(curr_data > 0.2)) / length(na.omit(curr_data)))
    })

    ## Median methylation
    df_cgi_summary_human$median_meth_prc2 <- apply(avg_cgi_human[rownames(avg_cgi_human) %in% prc2_cgis_human,samples_human], 2, median, na.rm = TRUE)
    df_cgi_summary_human$median_meth_pancancer <- apply(avg_cgi_human[rownames(avg_cgi_human) %in% hyper_pancancer_human,samples_human], 2, median, na.rm = TRUE)
    df_cgi_summary_human$median_meth_tumor_type <- sapply(samples_human, function(x)
    {
        curr_type <- df_cgi_summary_human[x,"type"]
        curr_data <- avg_cgi_human[rownames(avg_cgi_human) %in% hyper_cgi_per_tumor_type_array_any_purity[[curr_type]],x]
        return(median(curr_data, na.rm = TRUE))
    })
}

## Load and merge CGIs - mouse

df_cgi_summary_mouse <- data.frame(row.names = samples_mouse, type = sample_sheet[samples_mouse,"type"], condition = sample_sheet[samples_mouse,"condition"], group = sample_sheet[samples_mouse,"group"])

if (length(samples_mouse) > 0)
{
    avg_cgi_mouse <- lapply(samples_mouse, function(x)
    {
        data <- avg_cgi[[x]]
        colnames(data) <- c("chr", "start", "end", x)
        return(data)
    })

    avg_cgi_mouse <- Reduce(function(x,y) merge(x = x, y = y, all.x = TRUE, all.y = TRUE), avg_cgi_mouse)
    rownames(avg_cgi_mouse) <- paste(avg_cgi_mouse[,1], avg_cgi_mouse[,2], avg_cgi_mouse[,3], sep = "-")
    avg_cgi_mouse <- avg_cgi_mouse[,-c(1:3)]

    ## Fraction methylated
    df_cgi_summary_mouse$class <- "new"

    df_cgi_summary_mouse$fraction_meth_prc2 <- apply(avg_cgi_mouse[rownames(avg_cgi_mouse) %in% prc2_cgis_mouse,samples_mouse], 2, function(x) length(which(x > 0.2)) / length(na.omit(x)))
    df_cgi_summary_mouse$fraction_meth_pancancer <- apply(avg_cgi_mouse[rownames(avg_cgi_mouse) %in% hyper_pancancer_mouse,samples_mouse], 2, function(x) length(which(x > 0.2)) / length(na.omit(x)))
    df_cgi_summary_mouse$fraction_meth_tumor_type <- sapply(samples_mouse, function(x)
    {
        curr_type <- df_cgi_summary_mouse[x,"type"]
        curr_data <- avg_cgi_mouse[rownames(avg_cgi_mouse) %in% cgi_hg19_mm10_mapping[[curr_type]],x]
        return(length(which(curr_data > 0.2)) / length(na.omit(curr_data)))
    })

    ## Median methylation
    df_cgi_summary_mouse$median_meth_prc2 <- apply(avg_cgi_mouse[rownames(avg_cgi_mouse) %in% prc2_cgis_mouse,samples_mouse], 2, median, na.rm = TRUE)
    df_cgi_summary_mouse$median_meth_pancancer <- apply(avg_cgi_mouse[rownames(avg_cgi_mouse) %in% hyper_pancancer_mouse,samples_mouse], 2, median, na.rm = TRUE)
    df_cgi_summary_mouse$median_meth_tumor_type <- sapply(samples_mouse, function(x)
    {
        curr_type <- df_cgi_summary_mouse[x,"type"]
        curr_data <- avg_cgi_mouse[rownames(avg_cgi_mouse) %in% cgi_hg19_mm10_mapping[[curr_type]],x]
        return(median(curr_data, na.rm = TRUE))
    })
}

################################################################################
## Generate final data frames
################################################################################

## Columns of interest

columns_of_interest <- c("type", "condition", "group", "class", "fraction_meth_prc2", "median_meth_prc2", "fraction_meth_pancancer", "median_meth_pancancer", "fraction_meth_tumor_type", "median_meth_tumor_type")

## Combine information - human

metadata_human$class <- "reference"

df_cgi_summary_human_combined <- metadata_human[,columns_of_interest]
if (length(samples_human) > 0)
{
    df_cgi_summary_human_combined <- rbind(metadata_human[,columns_of_interest], df_cgi_summary_human[,columns_of_interest])
}
df_cgi_summary_human_combined_prc2 <- df_cgi_summary_human_combined[,c("type", "condition", "group", "class", "fraction_meth_prc2", "median_meth_prc2")]
df_cgi_summary_human_combined_pancancer <- df_cgi_summary_human_combined[,c("type", "condition", "group", "class", "fraction_meth_pancancer", "median_meth_pancancer")]
df_cgi_summary_human_combined_tumor_type <- df_cgi_summary_human_combined[,c("type", "condition", "group", "class", "fraction_meth_tumor_type", "median_meth_tumor_type")]

colnames(df_cgi_summary_human_combined_prc2) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")
colnames(df_cgi_summary_human_combined_pancancer) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")
colnames(df_cgi_summary_human_combined_tumor_type) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")

df_cgi_summary_human_combined_prc2$cgi <- "PRC2"
df_cgi_summary_human_combined_pancancer$cgi <- "Hyper pan-cancer"
df_cgi_summary_human_combined_tumor_type$cgi <- "Hyper tumor type"

df_cgi_summary_human_combined_stacked <- rbind(df_cgi_summary_human_combined_prc2, df_cgi_summary_human_combined_pancancer, df_cgi_summary_human_combined_tumor_type)
df_cgi_summary_human_combined_stacked$cgi <- factor(df_cgi_summary_human_combined_stacked$cgi, levels = c("PRC2", "Hyper pan-cancer", "Hyper tumor type"))
df_cgi_summary_human_combined_stacked$group <- factor(df_cgi_summary_human_combined_stacked$group, levels = c("primary", "model"))

## Combine information - mouse

metadata_mouse_models$class <- "reference"
metadata_mouse_models$group <- "model"

df_cgi_summary_mouse_combined <- metadata_mouse_models[,columns_of_interest]
if (length(samples_mouse) > 0)
{
    df_cgi_summary_mouse_combined <- rbind(metadata_mouse_models[,c(colnames(df_cgi_summary_mouse))], df_cgi_summary_mouse)
}

df_cgi_summary_mouse_combined_prc2 <- df_cgi_summary_mouse_combined[,c("type", "condition", "group", "class", "fraction_meth_prc2", "median_meth_prc2")]
df_cgi_summary_mouse_combined_pancancer <- df_cgi_summary_mouse_combined[,c("type", "condition", "group", "class", "fraction_meth_pancancer", "median_meth_pancancer")]
df_cgi_summary_mouse_combined_tumor_type <- df_cgi_summary_mouse_combined[,c("type", "condition", "group", "class", "fraction_meth_tumor_type", "median_meth_tumor_type")]

colnames(df_cgi_summary_mouse_combined_prc2) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")
colnames(df_cgi_summary_mouse_combined_pancancer) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")
colnames(df_cgi_summary_mouse_combined_tumor_type) <- c("type", "condition", "group", "class", "fraction_meth", "median_meth")

df_cgi_summary_mouse_combined_prc2$cgi <- "PRC2"
df_cgi_summary_mouse_combined_pancancer$cgi <- "Hyper pan-cancer"
df_cgi_summary_mouse_combined_tumor_type$cgi <- "Hyper tumor type"

df_cgi_summary_mouse_combined_stacked <- rbind(df_cgi_summary_mouse_combined_prc2, df_cgi_summary_mouse_combined_pancancer, df_cgi_summary_mouse_combined_tumor_type)
df_cgi_summary_mouse_combined_stacked$cgi <- factor(df_cgi_summary_mouse_combined_stacked$cgi, levels = c("PRC2", "Hyper pan-cancer", "Hyper tumor type"))
df_cgi_summary_mouse_combined_stacked$group <- factor(df_cgi_summary_mouse_combined_stacked$group, levels = c("primary", "model"))
```

## Human primaries and melanoma model - reference

```{r echo=FALSE, warning = FALSE, crop = NULL, fig.width = 10, fig.height = 3.5, message = FALSE, results='hide', fig.keep='all', fig.align = "center"}
ggplot(data = subset(df_cgi_summary_human_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, color = condition, shape = group)) + geom_point(size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_condition) + scale_shape_manual(values = shapes_group) + facet_grid(~cgi) 

ggplot(data = subset(df_cgi_summary_human_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, color = type, shape = condition)) + geom_point(size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_type) + scale_shape_manual(values = shapes_condition) + facet_grid(~cgi) 

```

```{=latex}
\newpage
```

`r  if (length(samples_human) > 0) {knitr::asis_output("## Human primaries and melanoma model - new\n")}`

```{r echo=FALSE, warning = FALSE, crop = NULL, fig.width = 10, fig.height = 3.5, message = FALSE, results='hide', fig.keep='all', fig.align = "center"}

if (length(samples_human) > 0)
{
    print(ggplot(data = subset(df_cgi_summary_human_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, shape = group)) + geom_point(color = "grey96", size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_condition) + scale_shape_manual(values = shapes_group) + facet_grid(~cgi) + geom_point(data = subset(df_cgi_summary_human_combined_stacked, class == "new"), aes(color = condition)))

    print(ggplot(data = subset(df_cgi_summary_human_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, shape = condition)) + geom_point(color = "grey96", size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_type) + scale_shape_manual(values = shapes_condition) + facet_grid(~cgi) + geom_point(data = subset(df_cgi_summary_human_combined_stacked, class == "new"), aes(color = type)))
}
```

```{=latex}
\newpage
```

## Mouse models - reference

```{r echo=FALSE, warning = FALSE, crop = NULL, fig.width = 10, fig.height = 3.5, message = FALSE, results='hide', fig.keep='all', fig.align = "center"}

## Reference data

ggplot(data = subset(df_cgi_summary_mouse_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, color = condition, shape = group)) + geom_point(size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_condition) + scale_shape_manual(values = shapes_group) + facet_grid(~cgi) 

ggplot(data = subset(df_cgi_summary_mouse_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, color = type, shape = condition)) + geom_point(size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_type) + scale_shape_manual(values = shapes_condition) + facet_grid(~cgi) 
```

```{=latex}
\newpage
```

`r  if (length(samples_mouse) > 0) {knitr::asis_output("## Mouse models - new\n")}`

```{r echo=FALSE, warning = FALSE, crop = NULL, fig.width = 10, fig.height = 3.5, message = FALSE, results='hide', fig.keep='all', fig.align = "center"}

if (length(samples_mouse) > 0)
{
    print(ggplot(data = subset(df_cgi_summary_mouse_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, shape = group)) + geom_point(color = "grey96", size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_condition) + scale_shape_manual(values = shapes_group) + facet_grid(~cgi) + ggtitle("Mouse models - new") + geom_point(data = subset(df_cgi_summary_mouse_combined_stacked, class == "new"), aes(color = condition)))

    print(ggplot(data = subset(df_cgi_summary_mouse_combined_stacked, class == "reference"), aes(x = median_meth, y = fraction_meth, shape = condition)) + geom_point(color = "grey96", size = 2) + theme_classic() + xlab("Median CGI methylation") + ylab("Fraction methylated CGIs") + theme(axis.text=element_text(size=10), axis.title=element_text(size=12)) + coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + scale_color_manual(values = colors_type) + scale_shape_manual(values = shapes_condition) + facet_grid(~cgi) + ggtitle("Mouse models - new") + geom_point(data = subset(df_cgi_summary_mouse_combined_stacked, class == "new"), aes(color = type)))

}

```